### 2024-06-28 paper/paper-papaja-revising.Rmd
revisions 6-24

---
title             : "Illusory Implications: Mere exposure to ideas can induce beliefs"
shorttitle        : "Illusory implications"

@@

abstract: |
  Numerous psychological findings have shown that mere exposure to ideas makes those ideas seem more true, a finding commonly referred to as the “illusory truth” effect.  Under predominant accounts of the illusory truth effect, initial exposure to a statement provides a metacognitive feeling of “fluency” or familiarity that, upon subsequent exposure, leads people to infer the statement is more likely to be true. However, genuine beliefs do not only affect truth judgments about individual statements, they also imply other beliefs and drive decision-making. Here, we consider whether exposure to “premise” statements affects people's truth ratings for novel “implied” statements, a pattern of findings we call the "illusory implication" effect. We argue these effects would constitute evidence for genuine belief change from mere exposure and  identify a handful of existing findings that offer preliminary support for this claim. Building upon these, we conduct three new pre-registered experiments to further test this hypothesis, finding additional evidence that exposure to “premise” statements affected participants’ truth ratings for novel “implied” statements, including for considerably more distant implications than those previously explored. Our findings suggest that the effects of mere exposure reach further than previously thought, with potentially consequential implications for concerns around mis- and disinformation.

@@

@@

mytable1 <- make_table_df(fit_imp_quiz_int) %>% 
  rename(`$CI_{2.5\\%}$` = `l-95% CI`,`$CI_{97.5\\%}$` = `u-95% CI`) %>%
  mutate(
    Term = case_when(
      Term == "B_implied_trueFimpliedT" ~ "Statement Truth",
      Term == "familiarity" ~ "Exposure",
      Term == "implication" ~ "Implication",
      # grepl("Intercept\\[", Term) ~ gsub("Intercept\\[(.*)\\]","$\\\\alpha_{\\1}$", Term),
      Term == "item_sd(Intercept)" ~ "$\\sigma_{item}$",
      Term == "subj_sd(Intercept)" ~ "$\\sigma_{subj}$",
      TRUE ~ Term
    )
  ) %>% 
  make_table("Population coefficients of Bayesian regression model for Implication effects in Experiment 1.")

mytable2 <- make_table_df(fit_imp_exp2) %>% 
  rename(`$CI_{2.5\\%}$` = `l-95% CI`,`$CI_{97.5\\%}$` = `u-95% CI`) %>%
  mutate(
    Term = case_when(
      Term == "B_implied_trueFimpliedT" ~ "Statement Truth",
      grepl("familiarity", Term) ~ "Exposure",
      Term == "implication" ~ "Implication",
      # grepl("Intercept\\[", Term) ~ gsub("Intercept\\[(.*)\\]","$\\\\alpha_{\\1}$", Term),
      # Term == "item_sd(Intercept)" ~ "$\\sigma_{item}$",
      # Term == "subj_sd(Intercept)" ~ "$\\sigma_{subj}$",
      TRUE ~ Term
    )
  ) %>% 
  make_table("Population coefficients of Bayesian regression model for Implication effects in Experiment 2.")

mytable3 <- make_table_df(fit_imp_maximal) %>% 
  rename(`$CI_{2.5\\%}$` = `l-95% CI`,`$CI_{97.5\\%}$` = `u-95% CI`) %>%
  mutate(
    Term = case_when(
      Term == "B_implied_trueFimpliedT" ~ "Statement Truth",
      Term == "familiarity" ~ "Exposure",
      Term == "implication" ~ "Implication",
      # grepl("Intercept\\[", Term) ~ gsub("Intercept\\[(.*)\\]","$\\\\alpha_{\\1}$", Term),
      grepl("familiarity", Term) ~ "Exposure",
      # Term == "item_sd(Intercept)" ~ "$\\sigma_{item}$",
      # Term == "subj_sd(Intercept)" ~ "$\\sigma_{subj}$",
      TRUE ~ Term
    )
  ) %>% 
  make_table("Population coefficients of Bayesian regression model for Implication effects in Experiment 3.")

@@

Every day, people are faced with a barrage of unsupported claims, from pestering ad campaigns to blatant disinformation on social media. Altogether, we live within an information environment that is more connected and more saturated than ever before in human history [@bak-coleman.etal2021]. If we are sufficiently critical consumers of media, can we benefit from this rich access to information without being exploited by advertisers or misled by misinformation? Perhaps not. Numerous psychological findings indicate that mere exposure to ideas makes those ideas appear more true, a finding commonly referred to as the “illusory truth” effect [@dekeersmaecker.etal2020; @dechene.etal2010; @fazio.etal2015; @hasher.etal1977; @pennycook.etal2018]. This effect suggests that we cannot exist in an environment of mis- and disinformation without being affected by it---without exposure to these ideas distorting our sense of what is true.

@@

The effect is one of "illusory" truth because it occurs following "mere exposure" to statements. That is, it occurs when the statements are seen or heard in a non-communicative context, such as when they are read during a true/false quiz. Generally, being told something---even by a source of unknown trustworthiness---is prima facie reason for believing it [@grice1989; also see @corneille.etal2020]. But reading a statement on a true/false quiz is not reason for believing it; the statement is just as likely to be false as to be true.

@@

Why should mere exposure affect later ratings of truth? Gilbert [-@gilbert1991] proposed that any idea entertained is by default first believed and must later be "unbelieved." However, predominant theoretical accounts attribute illusory truth effects not to belief but rather to metacognitive experiences of fluency or familiarity [@dechene.etal2010]. Research on metacognition has shown that people's behavior and judgments are not only influenced by the content of their thoughts, but also by the phenomenological experience of processing those thoughts [@alter.oppenheimer2009; @schwarz2015].

Metacognitive processing "fluency" is the experienced ease of processing or thinking about information [@alter.oppenheimer2009; @wanke.hansen2015]. When thinking is easy and fluent, people tend to judge the targets of that thinking more favorably than if thinking is challenging and disfluent [@wanke.etal1997]. A number of findings have shown that processing fluency contributes to the illusory truth effect: Mere prior exposure to statements makes later processing of those statements more fluent, and this metacognitive experience of fluency leads people to rate those statements as more true [@difonzo.etal2016; @fazio.etal2015; @unkelbach2007; @wang.etal2016].

@@

## Mere Exposure and Belief

There is something unsettling about the illusory truth effect and the lack of agency seems to imply about our own thinking. Even more worrisome, there could be important societal implications if merely being exposed to an idea causes people to adopt it as a belief: Pennycook and colleagues [-@pennycook.etal2018] argue that the illusory truth effect, combined with an environment of pervasive misinformation, has important consequences for the functioning of democratic society [also see @UnkelbachEtAl2019Truth]. The illusory truth effect suggests that mere exposure to misinformation could have impacts that cannot be stopped by fact-checking labels [@pennycook.etal2018], nor effectively curbed by retractions or corrective information [c.f. @ecker.etal2011; @lewandowsky.etal2012].

However, there are more to _beliefs_ than ratings of truth. Though a person’s assent to a proposition (whether they agree with or judge it to be truthful) is an important marker and convenient measure of belief, other features of belief are just as essential. For one, beliefs imply other beliefs: for instance, believing that “it is sunny out” implies the belief that “it is daytime”. And believing misinformation, such as “Dominion voting machines were rigged in the 2020 U.S. election” might imply the belief that “Donald Trump actually won the 2020 U.S. election.” Beliefs also inform action and decision-making, and whether or not someone holds a belief can be judged (at least partly) from their decisions. When truly believed, misinformation can have serious consequences. For instance, a recent study found that people who marked just one piece of vaccine misinformation as accurate were nearly twice as likely to be unvaccinated against COVID-19 compared to those who did not endorse any vaccine misinformation [@ognyanova.etal2021].

The extent to which the illusory truth effect operates entirely through metacognitive processes stands to determine the extent of its real-world consequences. Generally, it is the wider consequences of belief (beyond truth ratings for a specific proposition) that should be most concerning. If illusory truth effects operate only through metacognitive influences, then the impact of prior exposure depends on a subsequent re-exposure to a now fluent or familiar situation or statement (which might include one activating similar concepts). To be sure, this might have important consequences, such as increasing the tendency to share information online after having been exposed previously [@vellani.etal2023]. Still, the impact of metacognitive effects seem less likely to drive inferences or decision-making in more novel situations. Instead, their influences would be constrained to operating within contexts similar to the initial exposure.

@@

A crucial feature of belief is that beliefs are connected with other related beliefs. Thus, one way to test whether mere exposure affects beliefs is to examine whether mere exposure to statements can affect truth ratings for other related or implied statements. We refer to this hypothetical impact of mere exposure on implied statements as the "Illusory Implication" Effect.

We believe a number of existing findings provide at least preliminary evidence for this illusory implication effect. A number of researchers have found that mere exposure affects truth ratings for paraphrases and direct implications or contradictions [@arkes.etal1991; @bacon1979; @garcia-marques.etal2015; @silva.etal2017; @unkelbach.rom2017]. A crucial question is how distantly inferences are drawn following mere exposure, and whether or not metacognitive accounts can explain these findings.

Silva and colleagues [-@silva.etal2017] found that mere exposure to statements increased truth ratings for paraphrases of those statements that expressed the same idea while sharing only minimal surface-level linguistic features. Conversely, exposure to statements *decreased* truth ratings for rephrased contradictions of those statements. Silva and colleagues [-@silva.etal2017] explain their findings as resulting from conceptual fluency and disfluency produced by the initial mere exposure to statements: prior exposure "activates" the relevant concepts, easing or interfering with their retrieval and use in processing subsequent statements and thereby producing different metacognitive experiences.

However, findings from Unkelbach and Rom [-@unkelbach.rom2017] cast doubt on a purely fluency-based account of these types of findings. These researchers also found that exposure to statements decreased truth judgments for contradictions of those statements. Crucially however, they also found that responses were faster for contradictions of previously-seen statements than for novel statements, indicating that processing of these statements was *more* rather than less fluent. Thus "conceptual disfluency" is not likely to be an appropriate account of these findings [@unkelbach.rom2017]. Instead, they and other authors theorize that people tend to believe statements so long as they are generally coherent with respect to existing knowledge, even if they are presented in a context of mere exposure [@garcia-marques.etal2015; @gilbert1991; @unkelbach.rom2017].

@@

Prior work tested the effect of mere exposure on implied statements using only direct contradictions of the original statement (or paraphrases of direct contradictions). Here, we sought to further test for an "illusory implication effect" by testing the effect of mere exposure to statements on their more distant logical implications. To do so, we designed a series of studies to examine whether mere exposure to one statement (the “premise”) could affect truth ratings for another statement this premise would logically entail to be either true or false (the “implication”).

We developed materials involving more distant inferences that do not afford direct one-to-one mappings between shared concepts in the premise and implication statements. Thus, any reduction in truth ratings for the contradicted statements following exposure to the premise statements are not likely to be attributable to disfluency [cf. @silva.etal2017]. For instance, one statement pair was the premise statement, “No U.S. astronauts have died since the Challenger explosion in 1986” paired with the implied statement, “The space shuttle Columbia disintegrated over Texas in 2003” (true, but contradicted by the "premise"). In contrast with prior work examining conflicting statements, there is no direct one-to-one mapping of concepts between the two statements. Instead, a chain of inferences must be developed for a conclusion to be drawn: that the space shuttle Columbia contained astronauts, that its disintegration would have killed them, and that 2003 is later than 1986.

In addition, we make several additional contributions toward definitively demonstrating the illusory implication effect beyond existing studies by 1) testing the effect with large samples of participants (present studies $N = 1020$), 2) utilizing modern hierarchical statistical modeling methods, and 3) taking great pains to ensure that participants' truth ratings are not influenced by task demands (Experiments 2 and 3). Across our studies, we find further compelling evidence for this “illusory implication” effect, indicating that mere exposure has genuine impacts on beliefs extending beyond those attributable to metacognitive cues.

@@

All experiments were preregistered prior to data collection and all materials, data, and analysis code have been made publicly available on the Open Science Foundation website (https://osf.io/znq3y/). We report our approach to determining sample sizes, all data exclusions, all manipulations, and all measures in each study. Where appropriate, we explicitly differentiate between preregistered analyses and non-preregistered analyses in our results. All statistical models were fit using the `brms` R package [@burkner2017]. Our study was reviewed and approved by the Arizona State University Institutional Review Board under project identifier `STUDY00013322`. Participants provided informed consent before beginning the study.

@@

```{r fig1, fig.env="figure", fig.pos = "h", fig.align = "center", fig.cap="Average truth ratings for Implication statements in Experiment 1, broken down by exposure (exposed vs. unexposed), instructions (fact vs quiz conditions), implication type (false-implied-true vs true-implied-false). For visualization purposes, means were calculated by translating ordinal responses onto a scale from -2.5 to 2.5. Error bars indicate standard errors. "}
fig1 <- df %>% 
  filter(phase=="test") %>% 
  filter(item_type == "B") %>% 
  group_by(exposed, exposure_cond, B_implied_true) %>% 
  summarize(
    M = mean(resp_num),
    se = sd(resp_num)/sqrt(n()),
    ll = M - se,
    ul = M + se
  ) %>% 
  mutate(
    exposure_cond = if_else(exposure_cond=="fact", "Fact condition", "Quiz condition"),
    exposed = if_else(exposed==1, "Exposed", "Unexposed"),
    B_implied_true = if_else( B_implied_true == "T implied F", "Contradicted", "Entailed") #as.character(B_implied_true)
    ) %>% 
  ggplot(
    aes(x=B_implied_true, color=factor(exposed), shape=factor(exposed), y = M, ymin = ll, ymax=ul)
  ) +
  geom_pointrange(position=position_dodge(width=.5)) +
  # coord_flip() +
  scale_color_manual(values = c("#da1e28", "#0f62fe")) +
  labs(y = "Truth rating", x = "Implication type", color="Exposure", shape="Exposure") +
  facet_wrap(~exposure_cond) +
  theme_bw(base_size=9.5) +
  theme(panel.grid=element_blank(), legend.position="bottom") +
  theme(axis.text.x = element_text(angle = 10, vjust = 0.5, hjust=.5))

fig1
```

@@

Participants in the quiz condition were told that they were to be given a true/false quiz, so that some of the statements would be true and some false. These participants rated how confident they were that each of the presented statements were true or false. The purpose of the quiz condition was to provide "mere exposure" to these statements, to test for illusory truth and illusory implication effects.

@@

We tested for the illusory truth effect by comparing truth ratings for premise statements seen during exposure to those that had not been seen. Similarly, we tested for the illusory implication effect by comparing truth ratings for implication statements whose corresponding premise statements were and were not presented during the exposure phase. If mere exposure to "premise" statements affects participants' judgments of the "implication" statements, this would be evidence for an illusory implication effect.

@@

All models were fit using the `brms` R package [@burkner2017], with model posteriors estimated using the No-U-Turn Markov Chain Monte Carlo (MCMC) sampler implemented in Stan. Four MCMC chains were run for each model, with 2000 samples (1000 burn-in) drawn from each. Chains were assessed for convergence with $\hat{R}$ and the total estimated effective sample size was verified to be greater than 1000 for all parameters [@gelman.etal2014].

@@

Table 1 presents a summary of the posterior distribution estimated for the population-level coefficients in this model. The results indicate that both familiarity and the implication of the premise statements affected participants’ truth ratings. The implication effect is credibly greater than zero.

>>>>

---
title             : "Illusory Implications: Incidental exposure to ideas can induce beliefs"
shorttitle        : "Illusory implications"

@@

abstract: |
  Numerous psychological findings have shown that incidental exposure to ideas makes those ideas seem more true, a finding commonly referred to as the “illusory truth” effect.  Under predominant accounts of the illusory truth effect, initial exposure to a statement provides a metacognitive feeling of “fluency” or familiarity that, upon subsequent exposure, leads people to infer the statement is more likely to be true. However, genuine beliefs do not only affect truth judgments about individual statements, they also imply other beliefs and drive decision-making. Here, we consider whether exposure to “premise” statements affects people's truth ratings for novel “implied” statements, a pattern of findings we call the "illusory implication" effect. We argue these effects would constitute evidence for genuine belief change from incidental exposure and  identify a handful of existing findings that offer preliminary support for this claim. Building upon these, we conduct three new pre-registered experiments to further test this hypothesis, finding additional evidence that exposure to “premise” statements affected participants’ truth ratings for novel “implied” statements, including for considerably more distant implications than those previously explored. Our findings suggest that the effects of incidental exposure reach further than previously thought, with potentially consequential implications for concerns around mis- and disinformation.

@@

source("fig-code.R")

@@

mytable1 <- make_table_df(fit_imp_quiz_int) %>% 
  rename(`$CI_{2.5\\%}$` = `l-95% CI`,`$CI_{97.5\\%}$` = `u-95% CI`) %>%
  mutate(
    Term = case_when(
      Term == "B_implied_trueFimpliedT" ~ "Statement Truth",
      Term == "familiarity" ~ "Exposure",
      Term == "implication" ~ "Implication",
      # grepl("Intercept\\[", Term) ~ gsub("Intercept\\[(.*)\\]","$\\\\alpha_{\\1}$", Term),
      Term == "item_sd(Intercept)" ~ "$\\sigma_{item}$",
      Term == "subj_sd(Intercept)" ~ "$\\sigma_{subj}$",
      TRUE ~ Term
    )
  ) %>% 
  make_table("Population coefficients and 95\\% credible intervals of Bayesian regression model for Implication effects in Experiment 1.")

mytable2 <- make_table_df(fit_imp_exp2) %>% 
  rename(`$CI_{2.5\\%}$` = `l-95% CI`,`$CI_{97.5\\%}$` = `u-95% CI`) %>%
  mutate(
    Term = case_when(
      Term == "B_implied_trueFimpliedT" ~ "Statement Truth",
      grepl("familiarity", Term) ~ "Exposure",
      Term == "implication" ~ "Implication",
      # grepl("Intercept\\[", Term) ~ gsub("Intercept\\[(.*)\\]","$\\\\alpha_{\\1}$", Term),
      # Term == "item_sd(Intercept)" ~ "$\\sigma_{item}$",
      # Term == "subj_sd(Intercept)" ~ "$\\sigma_{subj}$",
      TRUE ~ Term
    )
  ) %>% 
  make_table("Population coefficients and 95\\% credible intervals of Bayesian regression model for Implication effects in Experiment 2.")

mytable3 <- make_table_df(fit_imp_maximal) %>% 
  rename(`$CI_{2.5\\%}$` = `l-95% CI`,`$CI_{97.5\\%}$` = `u-95% CI`) %>%
  mutate(
    Term = case_when(
      Term == "B_implied_trueFimpliedT" ~ "Statement Truth",
      Term == "familiarity" ~ "Exposure",
      Term == "implication" ~ "Implication",
      # grepl("Intercept\\[", Term) ~ gsub("Intercept\\[(.*)\\]","$\\\\alpha_{\\1}$", Term),
      grepl("familiarity", Term) ~ "Exposure",
      # Term == "item_sd(Intercept)" ~ "$\\sigma_{item}$",
      # Term == "subj_sd(Intercept)" ~ "$\\sigma_{subj}$",
      TRUE ~ Term
    )
  ) %>% 
  make_table("Population coefficients and 95\\% credible intervals of Bayesian regression model for Implication effects in Experiment 3.")

@@

Every day, people are faced with a barrage of unsupported claims, from pestering ad campaigns to blatant disinformation on social media. Altogether, we live within an information environment that is more connected and more saturated than ever before in human history [@bak-coleman.etal2021]. If we are sufficiently critical consumers of media, can we benefit from this rich access to information without being exploited by advertisers or misled by misinformation? Perhaps not. Numerous psychological findings indicate that incidental exposure to ideas makes those ideas appear more true, a finding commonly referred to as the “illusory truth” effect [@dekeersmaecker.etal2020; @dechene.etal2010; @fazio.etal2015; @hasher.etal1977; @pennycook.etal2018]. This effect suggests that we cannot exist in an environment of mis- and disinformation without being affected by it---without exposure to these ideas distorting our sense of what is true.

@@

The effect is one of "illusory" truth because it occurs following "incidental exposure" to statements. That is, it occurs when the statements are seen or heard in a non-communicative context, such as when they are read during a true/false quiz. Generally, being told something---even by a source of unknown trustworthiness---is prima facie reason for believing it [@grice1989; also see @corneille.etal2020]. But reading a statement on a true/false quiz is not reason for believing it; the statement is just as likely to be false as to be true.

@@

Why should incidental exposure affect later ratings of truth? Gilbert [-@gilbert1991] proposed that any idea entertained is by default first believed and must later be "unbelieved." However, predominant theoretical accounts attribute illusory truth effects not to belief but rather to metacognitive experiences of fluency or familiarity [@dechene.etal2010]. Research on metacognition has shown that people's behavior and judgments are not only influenced by the content of their thoughts, but also by the phenomenological experience of processing those thoughts [@alter.oppenheimer2009; @schwarz2015].

Metacognitive processing "fluency" is the experienced ease of processing or thinking about information [@alter.oppenheimer2009; @wanke.hansen2015]. When thinking is easy and fluent, people tend to judge the targets of that thinking more favorably than if thinking is challenging and disfluent [@wanke.etal1997]. A number of findings have shown that processing fluency contributes to the illusory truth effect: Incidental prior exposure to statements makes later processing of those statements more fluent, and this metacognitive experience of fluency leads people to rate those statements as more true [@difonzo.etal2016; @fazio.etal2015; @unkelbach2007; @wang.etal2016].

@@

## Incidental Exposure and Belief

There is something unsettling about the illusory truth effect and the lack of agency seems to imply about our own thinking. Even more worrisome, there could be important societal implications if merely being exposed to an idea causes people to adopt it as a belief: Pennycook and colleagues [-@pennycook.etal2018] argue that the illusory truth effect, combined with an environment of pervasive misinformation, has important consequences for the functioning of democratic society [also see @UnkelbachEtAl2019Truth]. The illusory truth effect suggests that incidental exposure to misinformation could have impacts that cannot be stopped by fact-checking labels [@pennycook.etal2018], nor effectively curbed by retractions or corrective information [c.f. @ecker.etal2011; @lewandowsky.etal2012].

However, there are more to _beliefs_ than ratings of truth. Though a person’s assent to a proposition (whether they agree with or judge statements expressing it to be truthful) is an important marker and convenient measure of belief, other features of belief are just as essential. For one, beliefs imply other beliefs: for instance, believing that “it is sunny out” implies the belief that “it is daytime”. And believing misinformation, such as “Dominion voting machines were rigged in the 2020 U.S. election” might imply the belief that “Donald Trump actually won the 2020 U.S. election.” Beliefs also inform action and decision-making, and whether or not someone holds a belief can be judged (at least partly) from their decisions. When truly believed, misinformation can have serious consequences. For instance, a recent study found that people who marked just one piece of vaccine misinformation as accurate were nearly twice as likely to be unvaccinated against COVID-19 compared to those who did not endorse any vaccine misinformation [@ognyanova.etal2021].

<!-- It is important to distinguish between statements and the propositions or ideas they express. Two different statements might express the same proposition: the pair of statements "I went to the store the day before yesterday" and "I went to the store two days ago" is one example. When incidental exposure produces metacognitive experiences like fluency, this might lead to increased endorsement of a specific statement, without truly affecting a person's attitude toward the proposition it represents. Instead, the effects of prior exposure would depend on features of subsequently-seen statements (e.g. lexical similarity: sharing words; or conceptual similarity: sharing concepts), rather than the propositions they express. -->

```{r theory-fig, echo = FALSE, message=FALSE, fig.align='center', fig.cap='Diagram depicting the theorized causal relations between latent beliefs, metacogntive experiences, and truth ratings for different types of statements. We highlight the paths from observable behavior which can only be traced back to exposure through beliefs in red.', out.width='0.75\\linewidth', fig.pos='h'}
knitr::include_graphics("theory-fig.pdf")
```

Figure \@ref(fig:theory-fig) depicts our hypotheses concerning the causal relationships between latent beliefs, metacognitive experiences, and overt behaviors such truth ratings and actions. As shown in the figure, experiences or manipulations that affect metacognitive experiences, such as incidental exposure, might produce increased truth ratings for statements expressing $P$ (either through lexical or conceptual similarity). However, beliefs should be expected to produce additional consequences. Here, we focus on inferences: a manipulation affecting belief in $P$ should increase truth ratings for statements expressing some proposition $Q$ where $P$ implies $Q$, and decrease belief in $R$ where $P$ implies $\neg R$. Figure \@ref(fig:theory-fig) also depicts a possible relation between metacognitive experiences and statements expressing $Q$. Where a statement $S_Q$ expressing $Q$ invokes a similar set of concepts as $S_P$, conceptual fluency might produce increased truth ratings.

Crucially, our theory draws no connection between metacognitive experiences and reduced truth judgments for statements expressing $R$ where $P$ implies $\neg R$. Such a connection would require a cogent theory of how conceptual "disfluency" might be engendered to reduce truth judgments without the process of inference that would implicate changes in belief. Current accounts have not developed such a theory [e.g. c.f. @AlterOppenheimer2009Uniting; @VogelEtAl2020Truth], and it appears unlikely to be workable: people frequently encounter chains of statements invoking different concepts, yet they do so seemingly without issue, as considering one topic does not necessarily make it more difficult to consider another unrelated topic.

We argue that where effects of incidental exposure on truth judgments are attributable to metacognitive processes, these effects do not generally constitute evidence for a change in beliefs. Applying the theoretical framework sketched in Figure \@ref(fig:theory-fig), cases where incidental exposure to a proposition $P$ increases truth ratings for statements $S_P$ expressing that proposition do not generally constitute evidence that incidental exposure has affected beliefs, since there is a causal pathway between exposure and this result that does not pass through belief. Likewise, evidence of exposure to $S_P$ increasing truth ratings for $S_Q$ where $S_Q$ shares a similar set of concepts with $S_P$ do not constitute evidence for effects on beliefs. However, evidence that exposure decreases truth ratings for $S_R$ or increases truth ratings for $S'_Q$ where $S'_Q$ does not directly share concepts with $S_P$ would constitute evidence for effects on beliefs.

Practically speaking, the extent to which effects of incidental exposure operate entirely through metacognitive processes stands to determine the extent of their real-world consequences. Generally, it is the wider consequences of belief (beyond truth ratings for a specific statement) that should be most concerning. If effects of incidental exposure operate only through metacognitive influences, then the impact of prior exposure depends on a subsequent re-exposure to a now fluent or familiar situation or statement (which might include one activating similar concepts). To be sure, metacognitive effects might have important consequences even if they do not affect beliefs, such as increasing the tendency to share information online after having been exposed previously [@vellani.etal2023]. Still, the impact of metacognitive effects seem less likely to drive inferences or decision-making in more novel situations. Instead, their influences would be constrained to operating within contexts similar to the initial exposure.

@@

A crucial feature of belief is that beliefs are connected with other related beliefs. Thus, one way to test whether incidental exposure affects beliefs is to examine whether incidental exposure to statements can affect truth ratings for other related or implied statements in ways that cannot be attributed to metacognitive effects. We refer to this hypothetical impact of incidental exposure on implied statements as the "Illusory Implication" Effect.

A key test of the illusory implication effect as opposed to other illusory truth effects is whether prior incidental exposure can _reduce_ judgments of truth. The experience of fluency is inherently relative: a metacognitive experience is only fluent or disfluent with respect to some point of comparison. For instance, utterances can be made relatively fluent by rhyming (McGlone & Tofighbakhsh, 2000) or they can be made relatively disfluent with the addition of speech markers like "um" and "uh" Brennan & Williams (1995). However, in the case of repetition-induced fluency (whether conceptual, lexical, or otherwise), repetition can produce fluency, but it is less clear how a such a manipulation might produce disfluency. Thus, it is difficult to see how effects of incidental exposure that reduce truth ratings can be explained by metacognitive factors.

A handful of existing findings provide some preliminary evidence for our hypothesized illusory implication effect. A number of researchers have found that incidental exposure affects truth ratings for novel statements contradicting previously exposed statements [@garcia-marques.etal2015; @silva.etal2017; @unkelbach.rom2017].  To our knowledge, three experiments have found evidence for this pattern of influence: Garcia-Marques (n = 58) [@garcia-marques.etal2015], Silva et al. [-@silva.etal2017] (Exp. 2, n = 45), and Unkelbach and Rom [-@unkelbach.rom2017] (Exp. 3; self-referent condition; n = 27). Such a pattern is not easily attributable to purely metacogntiive processes. Instead, several of these researchers [e.g., @garcia-marques.etal2015; @silva.etal2017] attribute these lowered truth ratings to the "meaning of the statements" rather than to conceptual fluency.

Garcia-marques and colleagues [-@garcia-marques.etal2015] induced incidental exposure by asking participants to rate how interesting statements were, warning them that half were true and half false. After this initial task inducing incidental exposure, a subsequent quiz featured a mix of novel statements, repeated statements, and statements contradicting those previously seen either immediately or after a 1-week delay. In the immediate-test condition, contradictory statements were rated as less true than novel statements.[^bacon]

[^bacon]: These researchers used materials following those used by Bacon [-@bacon1979, Exp. 2], who induced incidental exposure through a series of true-or-false quizzes. In this study, contradictory statements were rated as less true following a two-week delay, but only when participants were warned that the statements had been changed. This warning provides an avenue by which metacognitive cues could be used to inform truth judgments about contradictions without inducing beliefs.

Silva and colleagues [-@silva.etal2017] found that exposure to statements increased truth ratings for paraphrases of those statements that expressed the same proposition while sharing only minimal surface-level linguistic features. These researchers explained their findings as resulting from conceptual fluency produced by the initial exposure to statements: prior exposure "activates" the relevant concepts, easing with their retrieval and use in processing subsequent statements and thereby producing different metacognitive experiences. Silva and colleagues [-@silva.etal2017] also found that exposure to statements *decreased* truth ratings for rephrased contradictions of those statements. However, these findings must be qualified by the exposure context used here: Silva and colleagues asked participants to rate their interest in statements, but did not warn them that the statements could be true or false. Thus the pragmatics of the task are unclear, and participants might have believed these statements because they understood them to be presented as facts.

Unkelbach and Rom [-@unkelbach.rom2017] examined both implied and contradictory statements. They found that exposure to implied statements increased truth ratings (Exp 4.) and that statements saw a significant decrease in truth ratings following incidental exposure to their contradictions in a "self-relevant" processing condition (Exp. 4). Crucially, the results of this experiment appear to preclude any purely fluency-based account of reduced truth ratings following incidental exposure to contradictory statements: These researchers found that responses were faster for contradictions of previously-seen statements than for novel statements, indicating that processing of these statements was *more* rather than less fluent. Thus even if an appropriate notion of "conceptual disfluency" could be developed, it is unlikely it could be an appropriate account of these findings [@unkelbach.rom2017]. Instead, they and other authors theorize that people tend to believe statements so long as they are generally coherent with respect to existing knowledge, even if they are presented in a context of incidental exposure [@garcia-marques.etal2015; @gilbert1991; @silva.etal2017; @unkelbach.rom2017].

The presently-hypothesized illusory inference effect is a key theoretical link demonstrating the connection between incidental exposure and _beliefs_, and  all of the attendant downstream consequences associated with beliefs. A handful of prior findings suggest the existence of such an effect, yet this existing evidence seems far from dispositive, being based on only a handful of small studies with sometimes inconsistent methods.

@@

Prior work tested the effect of incidental exposure on implied statements using only direct contradictions of the original statement (or paraphrases of direct contradictions). Here, we sought to further test for an "illusory implication effect" by testing the effect of incidental exposure to statements on their more distant logical implications. To do so, we designed a series of studies to examine whether incidental exposure to one statement (the “premise”) could affect truth ratings for another statement this premise would logically entail to be either true or false (the “implication”).

In our studies, we sought to eliminate as much as possible the role of perceptual and conceptual fluency in affecting truth judgments. These efforts are aimed at demonstrating as clearly as possible a causal pathway  between incidental exposure and effects on truth ratings through changes in belief. That is, establishing a pathway from exposure to increased truth ratings for implications $Q$ and decreased truth ratings for contradictions $R$. Compared with prior work, we developed materials involving more distant inferences that do not afford direct one-to-one mappings between shared concepts in the premise and implication statements. Thus, any reduction in truth ratings for the contradicted statements following exposure to the premise statements are not reasonably attributable to disfluency (closing the alternate pathway depicted in Figure X). For instance, one statement pair was the premise statement, “No U.S. astronauts have died since the Challenger explosion in 1986” paired with the implied statement, “The space shuttle Columbia disintegrated over Texas in 2003” (true, but contradicted by the "premise"). In contrast with prior work examining conflicting statements, there is no direct one-to-one mapping of concepts between the two statements. Instead, a chain of inferences must be developed for a conclusion to be drawn: that the space shuttle Columbia contained astronauts, that its disintegration would have killed them, and that 2003 is later than 1986.

Additionally, we make several further contributions toward definitively demonstrating the illusory implication effect beyond existing studies by 1) testing the effect with large samples of participants (present studies $N = 1020$), 2) utilizing modern hierarchical statistical modeling methods to generalize beyond the exact items used, and 3) taking great pains to ensure that participants' truth ratings are not influenced by task demands (Experiments 2 and 3). Across our studies, we find further compelling evidence for this “illusory implication” effect, indicating that incidental exposure has genuine impacts on beliefs extending beyond those attributable to metacognitive cues.

@@

All experiments were preregistered prior to data collection and all materials, data, and analysis code have been made publicly available on the Open Science Foundation website (https://osf.io/znq3y/). In these preregistrations, we report our approach to determining sample sizes, all data exclusions, all manipulations, and all measures in each study. The analytic plan for Experiments 1 and 2 was preregistered in terms of its general designs, and a more detailed analytic plan was preregistred for Experiment 3. Where appropriate, we explicitly differentiate between preregistered analyses and non-preregistered analyses in our results. All statistical models were fit using the `brms` R package [@burkner2017]. Our study was reviewed and approved by the Arizona State University Institutional Review Board under project identifier `STUDY00013322`. Participants provided informed consent before beginning the study.

@@

```{r fig-exp1, fig.env="figure", fig.pos = "h", fig.width=6, fig.height = 3, fig.align = "center", fig.cap="Left: Test-phase responses and model estimates for Implication statements in Experiment 3, broken down by exposure (exposed vs. unexposed) and implication type (contradicted and entailed). Bars represent observed proportions. Points and errorbars indicate model predictions with 95% credible intervals. Right: Difference in proportions of responses between exposed and unexposed conditions at test, aggregated across true and false response options. Bars indicate observed differences, with points and errobars indicating model predictions and 50% credible intervals."}
library(tidybayes)

make_exp_figure(df_reg %>% filter(exposure_cond=="quiz"), fit_imp_quiz_int)

@@

Participants in the quiz condition were told that they were to be given a true/false quiz, so that some of the statements would be true and some false. These participants rated how confident they were that each of the presented statements were true or false. The purpose of the quiz condition was to provide "incidental exposure" to these statements, to test for illusory truth and illusory implication effects.

@@

We tested for the illusory truth effect by comparing truth ratings for premise statements seen during exposure to those that had not been seen. Similarly, we tested for the illusory implication effect by comparing truth ratings for implication statements whose corresponding premise statements were and were not presented during the exposure phase. If incidental exposure to "premise" statements affects participants' judgments of the "implication" statements, this would be evidence for an illusory implication effect.

@@

All models were fit using the `brms` R package [@burkner2017], with model posteriors estimated using the No-U-Turn Markov Chain Monte Carlo (MCMC) sampler implemented in Stan. Four MCMC chains were run for each model, with 2000 samples (1000 burn-in) drawn from each. MCMC diagnostics indicate good convergence: chains were assessed for convergence with $\hat{R}$ (all parameters $\hat{R} = 1 \pm .01$) and the total estimated effective sample size was verified to be greater than 1000 for all parameters [@gelman.etal2014]. All models were fit without divergences and without warnings for the Estimated Bayesian Fraction of Missing Information (eBFMI).

@@

Table 1 presents a summary of the posterior distribution estimated for the population-level coefficients in this model. The results indicate that both familiarity and the implication of the premise statements affected participants’ truth ratings. Figure 1 also visualizes the relevant posterior predictive distributions for this model's fixed effects, ignoring the random effects structure. The model does well in predicting the proportions of responses, with approximately accurate coverage as expected for the different sets of intervals. It is worth noting that these model estimates cannot be expected to perfectly correspond to the observed proportions, because the effects of individual items and participants have been (at least partially) accounted for in the random effects of the models.
